Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-08-10 tex/manuscript.tex
Update manuscript.tex
Corrected some minor things as per pdf coloring

Statisticians would most likely take extra precautions, should the sample or batch sizes be very small. However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain even as sample and batch sizes increase. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.

@@

The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner. Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies. The batch effect estimation errors will influence group effects in proportion to how well the group is represented in each batch:
\begin{equation}
\widetilde{Y}_{\textrm{-}j}
=\frac{\sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}}{n_{\textrm{-}j}}
=\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i),
\quad
n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}.
\end{equation}
In a balanced group--batch design, the effect is the same all groups and thus does not influence group comparisons. In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences. In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data. For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers. The documentation accompanying the \texttt{sva} library has a runnable example demonstrating how to adjust a data set with ComBat, followed by an F-test. Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.

@@

To explain more clearly what is happening, and quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.

@@

Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied? Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e. group labels not given as parameters to ComBat), problems related to preserving group differences will not occur. For other cases, a re-analysis without using this approach is the most rigorous path. However, such a re-analysis is not feasible if the downstream  analysis can not adjust for batch effects by itself. To reach a reliable result, batch effects need to be handled in some way or another. To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work. Even when this is available, the necessary statistics and bioinformatics skills and work hours could still be in short supply.

>>>>

Statisticians would most likely take extra precautions, should the sample or batch sizes be very small. However, the effects of batch adjustment using two-way ANOVA or ComBat on unbalanced data sets remain, even as sample and batch sizes increase. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.

@@

The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner. Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies. The batch effect estimation errors will influence group effects in proportion to how well the group is represented in each batch:
\begin{equation}
\widetilde{Y}_{\textrm{-}j}
=\frac{\sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}}{n_{\textrm{-}j}}
=\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i),
\quad
n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}.
\end{equation}
In a balanced group--batch design, the effect is the same for all groups and thus does not influence group comparisons. In an unbalanced design, however, the estimation errors $\hat\gamma_i-\gamma_i$ will induce systematic differences between groups which, when ignored in downstream analyses, may lead to over-confidence in estimated group differences. In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data. For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers. The documentation accompanying the \texttt{sva} library has a executable example demonstrating how to adjust a data set with ComBat, followed by an F-test. Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.

@@

To explain more clearly what is happening, and to quantify the size of the problem, we may consider the simple case of estimating the difference $\Delta\beta=\beta_A-\beta_B$ between two groups, $A$ and $B$, when there are $m$ batches with batch $i$ containing $n_{iA}$ and $n_{iB}$ samples from each of the two groups for $i=1,\ldots,m$.

@@

Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied? Essentially, when the batch--group configuration is balanced, or group difference is ignored  (i.e. group labels not given as parameters to ComBat), problems related to preserving group differences will not occur. For other cases, a re-analysis without using this approach is the most rigorous path. However, such a re-analysis is not feasible if the downstream  analysis can not independently adjust for batch effects. To reach a reliable result, batch effects need to be handled in some way or another. To make matters worse, a re-analysis relies on the availability of the raw data and a description of processing and analysis steps taken in the original work. Even when this is available, the necessary statistics and bioinformatics skills and work hours could still be in short supply.
